Inferring stellar rotation periods using Gaussian processes

### 2016-12-20 documents/ms.tex
tdm rewrite of section 3.3

\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
\usepackage{cite, natbib}
\usepackage{float}
\usepackage{epsfig}
\usepackage{cases}
\usepackage[section]{placeins}
\usepackage{graphicx, subfigure}
\usepackage{color}

@@

\altaffiltext{\columbia}{Simons Fellow, Department of Astronomy, Columbia
University, NY, NY, RuthAngus@gmail.com}
\altaffiltext{\oxford}{Subdepartment of Astrophysics, University of Oxford,
UK}
\altaffiltext{\uw}{Department of Astronomy, University of Washington, Seattle,
WA}
\altaffiltext{\princeton}{Department of Astronomy, Princeton University,
Princeton, NJ}

@@

In order to recover the rotation periods of the simulated light curves using
Gaussian processes, we sampled the posterior PDFs of the kernel
hyperparameters described in equation \ref{eq:QP}.
The likelihood function for a GP is similar to the simple Gaussian likelihood
function that is used for optimisation problems where the uncertainties are
Gaussian and uncorrelated.
The latter can be written
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\sum_{n=1}^N\frac{(y_n-\mu)^2}{\sigma_n^2}
    - \frac{N}{2}\ln(2\pi\sigma_n^2),
\end{equation}
\label{eq:chi2}
where $y_n$ are the data, $\mu$ is the mean model and $\sigma_n$ are the
Gaussian uncertainites on the data.
The equivalent equation in matrix notation is
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\bf{r}^T\bf{C}^{-1}\bf{r}-\ln|\bf{C}|
    + \mathrm{constant},
\end{equation}
\label{eq:lhf1}
where $\bf{r}$ is the vector of residuals and $\bf{C}$ is the covariance
matrix,
\begin{eqnarray}
	\mathbf{C} &=& \left (\begin{array}{cccc}
	\sigma^2_1 & \sigma_{2, 1} & \cdots & \sigma_{N, 1} \\
	\sigma_{1, 2} & \sigma^2_2 & \cdots & \sigma_{N, 2} \\
    && \vdots & \\
	\sigma_{1, N} & \sigma_{2, N} & \cdots & \sigma^2_N
\end{array}\right )
\end{eqnarray}
In the case where the uncertainties are uncorrelated, the noise is `white',
(which is a frequent assumption made by astronomers and is sometimes
justified) and the off-diagonal elements of the covariance matrix are zero.
However, in the case where there is evidence for correlated
`noise'\footnote{In our case the `noise' is actually the model!}, as in the
case of \Kepler\ light curves, those off-diagonal elements are non-zero.
With GP regression, a covariance matrix generated by the kernel function,
${\bf K}$ replaces ${\bf C}$ in the above equation.
Incidentally, this approach is the reverse of the regression techniques
usually employed by astronomers.
In most problems in astronomy one tries to infer the parameters that describe
the mean model and, if correlated noise is present, to marginalise over that
noise.
Here, the parameters describing the correlated noise are what we are
interested in and our mean model is simply a straight line at $y=0$.

One could either maximise this likelihood function in order to find the
best-fit hyper-parameters of the covariance kernel function or, as in our
approach, sample from the posterior PDFs of the hyper-parameters.
The advantage of the maximum-likelihood method is that the best-fit parameters
will be found much faster, but the uncertainities on the rotation period
will not be constrained.
In order to measure accurate uncertainties, the posterior PDFs of the
parameters must be explored.
Since obtaining accurate uncertainties on rotation periods is one of our main
motivations for this method, we use MCMC despite the added computational
expense.

We use the ACF period as an initial guess for the rotation period (this
decision is discussed further in section \textsection \ref{sec:discussion}).
We use Gausssian distributions in the natural logarithm of all the parameters
of our GP model.
The means and variances of the priors used were learned during the development
of this method.
In early stages of development we used very broad priors on all the parameters
uniform in log-space between -20 and 20.
We then examined the distributions of the model parameters for `sucessful' and
`unsuccessful' rotation period recoveries, where a recovery was termed
successful if it lay within 10\% of the injected period.
For the successful cases, the parameter distributions were narrow Gaussians.
The parameter distributions of the unsuccessful cases were broad Gaussians
with similar means.
We adopted the means and standard deviations of the narrow Gaussians
describing the parameter distributions of the successful cases to describe our
Gaussian priors.
The means and standard deviations of these $\ln$ Gaussian priors are given in
table \ref{tab:priors}.

\begin{table*}
\begin{center}
\caption{Means and standard deviations of the Gaussian priors on the natural
logarithms of the GP model parameters.}
\begin{tabular}{lcc}
Parameter & Mean & Standard Deviation \\
    \hline
    $\ln(A)$ & -13 & 2.7 \\
    $\ln(l)$ & 6.2 & 1.5 \\
    $\ln(\Gamma)$ & -1.4 & 1.5 \\
    $\ln$(Rotation period, $P_{rot}$) & $\ln$(ACF period) &
                2 $\times \ln$(ACF period) \\
    $\ln\sigma$ & -17 & 5 \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:priors}
The parameter positions were also initialised at the means of these
distributions during MCMC posterior exploration.
In addition, we set the following bounds on the rotation period and the $l$
parameter.
\begin{eqnarray}
 	&	\ln(0.5) < \ln(P_{\mathrm{rot}}) < \ln(T/2) \\ \nonumber
    &   0 < \ln(l) \\
\end{eqnarray}
\label{eq:priors}
Where $T$ is the total timespan of the light curve.
We limit rotation periods to be greater than half a day because we are chiefly
interested in rotation of FGK stars which are likely to be greater than this
lower limit, however this restriction could be relaxed in principle.
We also limit the rotation periods to be less than half of the total timespan
of the data since this is the longest period we expect to reliably recover.
We prevent $\ln(l)$ from falling beneath zero both because we expect that
astrophysical signals in light curves should have longer correlation
timescales than this and in practice we found that if $l$ is unrestricted, it
can occasionally wander towards a small value, resulting in an incorrect
rotation period estimate.

We subsampled the light curves in order to reduce computation time.
We randomly subsampled the data, retaining an average of four data points per
day.
This allowed us to recover periods down to half a day --- the minimum
timescale of interest for this project, whilst reducing computational costs.
In order to reduce computation time the light curves were split into small
sections and the likelihood for each section calculated individually, then
summed.
The matrix operations scale as $N\ln(N)^2$ where $N$ is the number of data
points in the light curve, however the likelihood calculation scales linearly
with the number of light curve sections used.
By adding together likelihoods of individial sections, the likelihood
evaluation scales as $mn\ln(n)^2$, where $n$ is the number of data points per
section and $m$ is the number of sections.
This method therefore significantly reduces computation time.
We split the simulated light curves into sections, 200 days long.
The {\tt george} \citep{George} python package was used to implement the GP
model which uses the fast matrix solver, HODLR \citep{Ambikasaran2014}.
Matrix operations performed with HODLR scale as $N \log ^2 (N)$, where $N$ is
the number of data points.
We used {\tt emcee} \citep{Foreman-Mackey2013}, an affine invariant
\citep{Goodman2010} ensemble sampler to explore the posterior PDFs of the
model parameters.
The rotation period was taken as the sample with the maximum likelihood value.
The resulting measured rotation periods are compared to the true rotation
periods in figure \ref{fig:compare_mcmc}.

>>>>

\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
\usepackage{cite, natbib}
\usepackage{float}
\usepackage{epsfig}
\usepackage{cases}
\usepackage[section]{placeins}
\usepackage{graphicx, subfigure}
\usepackage{color}
\usepackage{bm}

@@

\altaffiltext{\columbia}{Simons Fellow, Department of Astronomy, Columbia
University, NY, NY, RuthAngus@gmail.com}
\altaffiltext{\oxford}{Subdepartment of Astrophysics, University of Oxford,
UK}
\altaffiltext{\uw}{Department of Astronomy, University of Washington, Seattle,
WA}
\altaffiltext{\princeton}{Department of Astrophysical Sciences, Princeton University,
Princeton, NJ}

@@

In order to recover the rotation period of a simulated light curve using
Gaussian processes, we sample the following posterior PDF:
\begin{equation}
p({\bm \theta}\,|\,y) \propto \mathcal L(y\,|\,{\bm \theta}) p({\bm \theta}),
\end{equation}
\label{eq:posterior}
where $y$ are the light curve data, $\bm \theta$ are the hyperparameters 
of the kernel described in equation \ref{eq:QP}, $\mathcal L$ is the 
GP likelihood function, and $p({\bm \theta})$ is the prior on the 
hyperparameters.  The GP likelihood is similar to the simple Gaussian likelihood
function used for optimisation problems where the uncertainties are
Gaussian and uncorrelated. The latter can be written
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\sum_{n=1}^N\frac{(y_n-\mu)^2}{\sigma_n^2}
    - \frac{N}{2}\ln(2\pi\sigma_n^2),
\end{equation}
\label{eq:chi2}
where $y_n$ are the data, $\mu$ is the mean model and $\sigma_n$ are the
Gaussian uncertainites on the data.
The equivalent equation in matrix notation is
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\bf{r}^T\bf{C}^{-1}\bf{r}-\ln|\bf{C}|
    + \mathrm{constant},
\end{equation}
\label{eq:lhf1}
where $\bf{r}$ is the vector of residuals and $\bf{C}$ is the covariance
matrix,
\begin{eqnarray}
	\mathbf{C} &=& \left (\begin{array}{cccc}
	\sigma^2_1 & \sigma_{2, 1} & \cdots & \sigma_{N, 1} \\
	\sigma_{1, 2} & \sigma^2_2 & \cdots & \sigma_{N, 2} \\
    && \vdots & \\
	\sigma_{1, N} & \sigma_{2, N} & \cdots & \sigma^2_N
\end{array}\right )
\end{eqnarray}
In the case where the uncertainties are uncorrelated, the noise is `white',
(which is a frequent assumption made by astronomers and is sometimes
justified) and the off-diagonal elements of the covariance matrix are zero.
However, in the case where there is evidence for correlated
`noise'\footnote{In our case the `noise' is actually the model!}, as in the
case of \Kepler\ light curves, those off-diagonal elements are non-zero.
With GP regression, a covariance matrix generated by the kernel function,
${\bf K}$ replaces ${\bf C}$ in the above equation.
Incidentally, this approach is the reverse of the regression techniques
usually employed by astronomers.
In most problems in astronomy one tries to infer the parameters that describe
the mean model and, if correlated noise is present, to marginalise over that
noise.
Here, the parameters describing the correlated noise are what we are
interested in and our mean model is simply a straight line at $y=0$.

Sampling the posterior described above presents several challenges:
\begin{itemize}
\item Evaluating $\mathcal L$ for an entire \Kepler\ lightcurve 
($\sim$40,000 points) takes about $\sim$5\,s--- too computationally expensive 
to perform inference on large numbers of light curves\footnote{All computational
times cited in this section are based on evaluations on a 
single core of a 2015 Macbook Pro, 3.1 GHz Intel Core i7.}.  The matrix operations
necessary to evaluate the GP likelihood (using the fast matrix solver 
HODLR \citep{Ambikasaran2014}, implemented in the {\tt george} \citep{George} 
python package) scale as $N\ln(N)^2$, where $N$ is
the number of data points in the light curve.

\item The flexibility of this GP model allows for insidious kinds of 
posterior multimodality and "over-fitting"--like behavior.  For example, 
if $l$ is small,  the non-periodic factor in the covariance kernel may dominate, 
allowing for a good fit to the data without requiring any 
periodic covariance structure---even if
clear periodic structure is present.  Additionally, if $\Gamma$ is too large,
then [explain in more technical terms that "the GP gets way too wiggly and
overfits the data without caring about the periodicity." ]

\item In some cases, the posterior may also be multi-modal in period---that is, 
there may be significant posterior density at half or twice the true period $P$.
This is related to how the ACF method must guard against identifying periods 
at half or twice the true period. 
\end{itemize}

We address the first of these difficulties in two distinct ways: subsampling
the data and splitting the light curve into independent sections.  
To subsample, we randomly select 1/30th of the points in the full light curve (an 
average of $\sim$1.5 points per day).  This decreases the likelihood evaluation 
time by a factor of about 50, down to about 100\,ms.
We then split the light curve into equal-sized
chunks containing approximately 300 points per section (corresponding to about
200 days), and evaluate the log-likelihood as the sum of the log-likelihoods
of the individual sections (all using the same parameters $\theta$).  This reduces
computation time because the section-based likelihood evaluation scales as $mn\ln(n)^2$, where $n$ is the number of data points per section and $m$ is the number of sections.
This method further reduces computation time for a typical light curve (subsampled
by a factor of 30) by about a factor of two, to about 50\,ms.

In order to manage the extreme flexibilty of the GP model and focus on 
reliably retrieving the period parameter, we also impose priors
on the non-period GP parameters that we 
learned during the development of this method.  
Beginning with very broad priors (uniform in log space between 
-20 and 20) on all the parameters, we first explored what the 
typical values of the non-period parameters tended to be when the 
injected periods were successfully recovered.  We also experimented with
constraining the allowed ranges of the parameters, after discovering 
that some regions of parameter space (such as large values of $A$ and $\Gamma$
and small values of $l$) tended to allow fits that ignored the desired
periodicity.  The priors we used for this work are listed in
table \ref{tab:priors}.  
% We note that our default prior for period is 
% log-uniform between the bounds, but that we were also able to 
% further improve our results by using an ACF-inspired prior.

To sample the posterior in a way that is sensitive to potential 
multimodality, we use the \texttt{emcee3} MCMC sampler.
We initialize 500 walkers with random samples from the prior 
 and use a mix of different proposal distributions: 
specifically, a 2:2:1 mix of KDE, differential evolution, and DESnooker moves. 
[DFM: more words/citations here?]  
We run fifty steps of the sampler at a time, checking for convergence after 
each iteration, up to a maximum of fifty iterations.  We declare convergence if 
the total effective chain length is at least 8$\times$ the maximum autocorrelation
time.  When convergence is achieved, we drop the first two autocorrelation 
lengths in the chain as a burn-in, and randomly choose 5000 samples as 
representative of the posterior.  This fitting process takes several hours 
for a typical simulated light curve.

Figure \ref{fig:compare_mcmc} summarizes the results of this MCMC fitting procedure
compared to the injected ``true'' stellar rotation periods.

\begin{table*}
\begin{center}
\caption{Priors and bounds on the natural logarithms of the GP model parameters.}
\begin{tabular}{lcc}
Parameter & Prior & Bounds\\
    \hline
    $\ln A$ & $\mathcal N(-13, 5.7)$ & (-20, 0) \\
    $\ln l$ & $\mathcal N(7.2, 1.2)$ & (2, 20) \\
    $\ln \Gamma$ & $\mathcal N(-2.3, 1.4)$ & (-10, 3) \\
    $\ln \sigma$ & $\mathcal N(-17, 5)$ & (-20, 0) \\
    $\ln P $ & Uniform & ($\ln 0.5, \ln 100$) \\ 
\end{tabular}
\end{center}
\end{table*}
\label{tab:priors}
